---
title: "Reproducible Research Project 1 (Activity Data)"
author: "Janet Miceli"
date: "October 10, 2017"
output: html_document
---

##Loading and preprocessing the data
1.Load the data

```{r}
url="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
library(downloader)
download(url, dest="activity.zip", mode="wb")
unzip("activity.zip", exdir=".")
activity <- read.csv("activity.csv")
```

2.Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
activity$date <-as.Date(activity$date, format="%Y-%m-%d")
```
##What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.


1.Make a histogram of the total number of steps taken each day

```{r}    
StepsPerD <-aggregate(steps ~ date, data= activity, sum, na.rm = TRUE)
par(mfrow=c(1, 1))

hist(StepsPerD$steps, breaks=20,
     main="Total Number of Steps Taken per Day", xlab= "Steps")
```
2.Calculate and report the mean and median total number of steps taken per day

```{r}
mean(StepsPerD$steps)

median(StepsPerD$steps)
```

##What is the average daily activity pattern?
1.Make a time series plot (i.e.  type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
av_step <-aggregate(steps ~ interval, data = activity, mean, na.rm = TRUE)
plot(av_step$interval, av_step$steps, type = "l", lwd=2,
     main="Time Series, Average Number of Steps per Interval Across All Days", axes = FALSE,
     xlab = " 5-minute interval No.", ylab="Average No. of Steps")
```
2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
av_step$interval[which.max(av_step$steps)]
```
The 835th 5-minute interval containes the maximum number of steps.

##Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as  NA ). The presence of missing days may introduce bias into some calculations or summaries of the data.

1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with  NA s)

```{r}
sum(is.na(activity))
```
There are 2304 missing values in this dataset.

2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

In order to fill in missing values in this data set, the mean  number of steps for that five-minute interval will be used

3.Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
impute <- activity
for(i in av_step$interval){
    impute[impute$interval == i & is.na(impute$steps), ]$steps <-
        av_step$steps[av_step$interval==i]
}
head(impute)
```
4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
StepsPerDI <-aggregate(steps ~ date, data = impute, sum, na.rm = TRUE)
hist(StepsPerDI$steps, breaks = 20, 
     main = "Total Number of Steps Taken per Day(imputed)",
     xlab="Steps")

mean(StepsPerDI$steps)

median(StepsPerDI$steps)
```
Imputing had no impact on the mean, however the median is slightly influences. Imputing using the average of the five minute interval results in more data representing the mean, leading to less variation.

##Are there differences in activity patterns between weekdays and weekends?

"For this part the  weekdays()  function may be of some help here. Use the dataset with the filled-in missing values for this part."

1.Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
impute$day <-weekdays(impute$date)
impute$week <- ""
impute[impute$day == "Saturday" | impute$day == "Sunday", ]$week <- "weekend"
impute[!impute$day == "Saturday" | impute$day == "Sunday", ]$week <- "weekday"
impute$week <-factor(impute$week)
```

2.Make a panel plot containing a time series plot (i.e.  type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using simulated data:

```{r}
av_stepI <- aggregate(steps ~ interval + week, data = impute, mean)
library(lattice)
xyplot(steps ~ interval | week, data = av_stepI, type = "l", lwd = 2, 
       layout=c(1, 2),
       xlab="5-minute Interval",
       ylab= "Average Number of Steps",
       main= "Average No. of Steps (across all weekday or weekend days)")
```

